NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PE and PG lipids

### 2020-11-17 Manuscript/manuscriptPGPE.tex
Update manuscript

\clearpage
\subsection{Lipid conformational ensembles in lipid bilayers with bound ions}

Headgroup order parameters of PC lipids decrease or increase upon addition of positive
or negative charge into membrane, respectively, while they remain unchanged upon addition
of neutral molecules such as cholesterol or sphingomyelin \cite{??}.
This has been explained by the tilting of the headgroup dipole due to electrostatic
interactions with bound charges, and has been used to measure the binding affinity
of ions to membranes \cite{??}.
%The headgroup order parameters of PC lipids can be used to measure the ion binding
%affinity to lipid bilayers because their magnitude is linearly proportional to the amount of bound charge in bilayer \cite{seelig87,catte16}.
%This molecular electrometer concept can be used also for bilayers containing PC lipids mixed with charged lipids \cite{borle85,macdonald87,roux90,antila19}.
%The headgroup order parameters can be used to evaluate MD simulations against experimental data,
%because they can be directly calculated from MD simulations \cite{catte16}.

%To evaluate the calcium binding in simulations of lipid bilayers containing PG lipids,
%we calculated
The changes of PC headgroup order parameters in POPC:POPG (1:1) and (4:1) mixtures
upon addition of CaCl$_2$ from simulations and experiments \cite{borle85,macdonald87}
%, and compared these with the available experimental data \cite{borle85,macdonald87}.
in Figs. \ref{changesWITHCaClPG} and \ref{changesWITHCaClPG1PC4} are in line with
conclusion from our previous studies \cite{??}:  
%The decrease of POPC headgroup order parameters in mixtures with POPG lipids with
%increasing CaCl$_2$ concentration is overestimated in Slipids and Lipid17
order parameters decrease in the response to bound cations in qualitative agreement with
experiments, but calcium binding affinity to membranes is overestimated in 
simulations except in CHARMM36 with the NBfix correction which underestimates the binding affinity.
On the other hand, implicit inclusion of electronic polarizability to Amber based lipid models
improves the ion binding affinity leading to a good agreement with experiments.
%(Figs. \ref{changesWITHCaClPG} and \ref{changesWITHCaClPG1PC4})
%
%indicating too strong binding affinity of calcium into the bilayers as previously
%observed for pure PC lipid bilayers and mixtures with PS lipids \cite{catte16,antila19}.
%
%The decrease of $\alpha$-carbon order parameter of PC lipids in PC:PG mixtures as a function
%of calcium concentration is close to experiments CHARMM36 simulations (Fig. \ref{changesWITHCaClPG}),
%but the decrease of $\beta$-carbon order parameter seems to be overestimated.
%However, the $\beta$-carbon order parameter was not actually measured from these samples,
%but they are calculated from empirical relation $\Delta S_{\beta}=0.43\Delta S_{\alpha}$ \cite{akutsu81}.
%On the other hand, the simulations are not converged.
%
%The calcium binding affinity to lipid bilayers with PC and PS lipids
%was recently improved by applying the electronic continuum correction (ECC) to Amber Lipid14/17 force fields \cite{melcr18,melcr20}.
%In this approach, the electronic polarizability is implicitly included in the classical force fields
%by scaling the charges with constant factors \cite{leontyev11}. Here, we make a ECC-POPG force field
%by applying the scaling factors originally used for POPS also to POPG, i.e., we multiply
%charges and Lennard-Jones $\sigma$s of headgroup, glycerol backbone, and carbonyl regions with 
%$f_q$=0.75 and $f_\sigma$=0.89, respectively \cite{melcr20}.
%ECC-POPG model gives a weaker calcium binding affinity (Fig.\ref{CAdensPG}) and better agreement with the experimental
%PC headgroup order parameter changes (Fig. \ref{changesWITHCaClPG}) for POPC:POPG mixtures than the original Lipid17 model
%\todo{to be finished when we have all the data}, indicating that the ECC improves the simulation predictions of
%calcium binding affinity as previously observed for PC and PS lipids \cite{melcr18,melcr20}.

>>>>

\clearpage
\subsection{Lipid conformational ensembles in lipid bilayers with bound ions}
Phosphatidylcholine headgroup conformational ensemble and orientation are unchanged
upon addition of neutral molecules such as cholesterol or sphingomyelin into a bilayer,
while incorporation of charges affect the headgroup dipole vector tilt which can
be detected by measuring C-H bond order parameters of $\alpha$ and $\beta$ carbons \cite{??}.
Such changes have been observed upon addition of charged lipids, proteins, surfactants,
drugs, or ions \cite{??}, but how charges affect the lipid conformational ensemble remains unknown.
%Headgroup order parameters of PC lipids decrease or increase upon addition of positive
%or negative charge into membrane, respectively, while they remain 
%\cite{??}.
%This has been explained by the tilting of the headgroup dipole due to electrostatic
%interactions with bound charges, and has been used to measure the binding affinity
%of ions to membranes \cite{??}.
%The headgroup order parameters of PC lipids can be used to measure the ion binding
%affinity to lipid bilayers because their magnitude is linearly proportional to the amount of bound charge in bilayer \cite{seelig87,catte16}.
%This molecular electrometer concept can be used also for bilayers containing PC lipids mixed with charged lipids \cite{borle85,macdonald87,roux90,antila19}.
%The headgroup order parameters can be used to evaluate MD simulations against experimental data,
%because they can be directly calculated from MD simulations \cite{catte16}.

\begin{figure*}[bt]
  \centering
  \includegraphics[width=9.0cm]{../Figs/HGorderparametersPCvsSURFchange.eps}
  \caption{\label{changesWITHsurf}
    Modulation of PC headgroup order parameters and P-N vector angle upon addition of cationic surfactant.
  }
\end{figure*}

\begin{figure}[]
  \centering
  \includegraphics[width=6.0cm]{../Figs/dihedralFIGS/POPCsurf_M_G3O3_M_M_G3C4_M_M_G3C5_M_M_G3N6_MCHARMM36.png}
  \includegraphics[width=6.0cm]{../Figs/dihedralFIGS/POPCsurf_M_G3P2_M_M_G3O3_M_M_G3C4_M_M_G3C5_MCHARMM36.png}
  \includegraphics[width=6.0cm]{../Figs/dihedralFIGS/POPCsurf_M_G3O1_M_M_G3P2_M_M_G3O3_M_M_G3C4_MCHARMM36.png}
  \includegraphics[width=6.0cm]{../Figs/dihedralFIGS/POPCsurf_M_G3_M_M_G3O1_M_M_G3P2_M_M_G3O3_MCHARMM36.png}
  \includegraphics[width=6.0cm]{../Figs/dihedralFIGS/POPCsurf_M_G2_M_M_G3_M_M_G3O1_M_M_G3P2_MCHARMM36.png}
  \includegraphics[width=6.0cm]{../Figs/dihedralFIGS/POPCsurf_M_G1_M_M_G2_M_M_G3_M_M_G3O1_MCHARMM36.png}  
  \caption{\label{DIHSwithSURF}
    Changes in CHARMM36 dihedrals with increasing amount of cationic surfactant.
  }
\end{figure}

The response of PC headgroup order parameters to the addition of cationic surfactants into a
membrane in CHARMM36 simulations are in very good agreement with the experimental data (Fig. \ref{changesWITHsurf})
suggesting that these simulations can be used to interpret how lipid headgroups respond to the
inclusion of charges into a membrane. Analysis of lipid dihedral bond distributions suggest that
the lipid conformational changes upon addition of charge occur in phosphate-oxygen and g$_3$-phosphate-oxygen bonds.

To resolve the lipid headgroup conformational ensembles in mixed membranes we analyzed
headgroup order parameters from mixture of PC and PE or PG and compared the changes
upon mixing to the experimental data \cite{??} (Figs. \ref{??}). However, the currently
available force fields failed to capture the headgroup conformational ensembles in mixed membranes
as we previously observed also for mixtures with PS lipids \cite{antila19}. The best performing force field for
single componen membrane, CHARMM36, overestimated the influcence of PE to PC conformations and
undersestimated response to anionic PG headgroup. The latter may arise from overbinding of sodium
counterions \cite{antila19}.

Also the changes of headgroup order parameters in POPC:POPG (1:1) and (4:1) mixtures
upon addition of CaCl$_2$
%from simulations and experiments \cite{borle85,macdonald87}
in Figs. \ref{changesWITHCaClPG} and \ref{changesWITHCaClPG1PC4} are in line with
conclusions from our previous studies \cite{catte16,melcr18,antila19,melcr20}:  
%The decrease of POPC headgroup order parameters in mixtures with POPG lipids with
%increasing CaCl$_2$ concentration is overestimated in Slipids and Lipid17
%order parameters decrease in the response to bound cations in qualitative agreement with
%experiments, but
calcium binding affinity to membranes is typically overestimated in simulations, but
CHARMM36 with the NBfix correction underestimates the binding affinity, and that the
%On the other hand,
results can be improved by implicit inclusion of electronic polarizability.
Among the available simulations, there are two models that correctly reproduce the
PG headgroup order parameter changes upon addition of calcium, Lipid17 and Slipids,
even though the binding affinity of calcium is overestimated in these simulations.
The response of PC to calcium binding is correctly reproduced by Lipid17ecc model, as
previously reported for other membranes \cite{melcr18,melcr19}.
%to Amber based lipid models
%improves the ion binding affinity leading to a good agreement with experiments.
%(Figs. \ref{changesWITHCaClPG} and \ref{changesWITHCaClPG1PC4})
%
%indicating too strong binding affinity of calcium into the bilayers as previously
%observed for pure PC lipid bilayers and mixtures with PS lipids \cite{catte16,antila19}.
%
%The decrease of $\alpha$-carbon order parameter of PC lipids in PC:PG mixtures as a function
%of calcium concentration is close to experiments CHARMM36 simulations (Fig. \ref{changesWITHCaClPG}),
%but the decrease of $\beta$-carbon order parameter seems to be overestimated.
%However, the $\beta$-carbon order parameter was not actually measured from these samples,
%but they are calculated from empirical relation $\Delta S_{\beta}=0.43\Delta S_{\alpha}$ \cite{akutsu81}.
%On the other hand, the simulations are not converged.
%
%The calcium binding affinity to lipid bilayers with PC and PS lipids
%was recently improved by applying the electronic continuum correction (ECC) to Amber Lipid14/17 force fields \cite{melcr18,melcr20}.
%In this approach, the electronic polarizability is implicitly included in the classical force fields
%by scaling the charges with constant factors \cite{leontyev11}. Here, we make a ECC-POPG force field
%by applying the scaling factors originally used for POPS also to POPG, i.e., we multiply
%charges and Lennard-Jones $\sigma$s of headgroup, glycerol backbone, and carbonyl regions with 
%$f_q$=0.75 and $f_\sigma$=0.89, respectively \cite{melcr20}.
%ECC-POPG model gives a weaker calcium binding affinity (Fig.\ref{CAdensPG}) and better agreement with the experimental
%PC headgroup order parameter changes (Fig. \ref{changesWITHCaClPG}) for POPC:POPG mixtures than the original Lipid17 model
%\todo{to be finished when we have all the data}, indicating that the ECC improves the simulation predictions of
%calcium binding affinity as previously observed for PC and PS lipids \cite{melcr18,melcr20}.
